---
title: "Texas_Blackout_analysis"
author: "Maeve Li Oak"
date: '11/7/2025'
format: html
editor: visual
---

### Loading Packages

This portion of the document loads all the packages necessary to complete the rest of the document. The goal of this project is to analyze the impact of climate change induced extreme weather events through an analysis of the 2021 Texas blackouts. The data for this project will be focused on remotely sensed night lights data as a identifier for areas which lost power.

```{r}
#loading packages
#| message: false
#| warning: false


library(sf) # vector handling
library(terra) # raster handling
library(tidyverse)
library(tmap) # map making
library(spData) # spatial data
library(spDataLarge) # spatial data
library(viridisLite)
library(raster)
library(ggplot2)
library(mosaic)
library(stars)
library(lwgeom)

```

# Loading Data

```{r}
#loading data 
#| message: false
#| warning: false

#loading geopackages for roads, houses 
roads <- st_read(here::here("data", "gis_osm_roads_free_1.gpkg"),
                 query = "SELECT * FROM gis_osm_roads_free_1 WHERE fclass='motorway'",
                 quiet = TRUE) %>%
                 st_make_valid()

houses <- st_read(here::here("data", "gis_osm_buildings_a_free_1.gpkg"), 
                  query = "SELECT *
                  FROM gis_osm_buildings_a_free_1
                  WHERE(type IS NULL AND name IS NULL)
                  OR type in ('residential', 'apartments', 'house', 'static_caravan', 'detached')",
                  quiet = TRUE) %>%
                  st_make_valid()

#loading Socioeconomic data
#commented out is loading the dataframe entirely, and reviewing the names of the different layers in order to pull the relevant information 

#st_layers(here::here(data, ACS_2019_5YR_TRACT_48_TEXAS.gdb))
#SES <- st_read(here::here("data", "ACS_2019_5YR_TRACT_48_TEXAS.gdb")) #returning a dataframe, it's non spatial which is ok! 
#pulling the geometry from the dataset
geo <- st_read(here::here('data', 'ACS_2019_5YR_TRACT_48_TEXAS.gdb'),
               layer = 'ACS_2019_5YR_TRACT_48_TEXAS',
               quiet = TRUE) %>%
               st_make_valid() 
#pulling the income variable 
income <- st_read(here::here('data', 'ACS_2019_5YR_TRACT_48_TEXAS.gdb'),
               layer = 'x19_income',
               quiet = TRUE) 
#extracting the metadata
meta <- st_read(here::here('data', 'ACS_2019_5YR_TRACT_48_TEXAS.gdb'),
               layer = 'TRACT_METADATA_2019',
               quiet = TRUE) 

```

### Creating a blackout mask

This portion creates a mask which indicates whether or not a Texas cell experienced a blackout.

In order to do this properly I first need to find the change in night lights caused by the storm, which requires me to make a raster object for each day, this means combining light data from two areas (area 05, and area 06) of Texas into a raster for each day (Feb 7th and 16th).

```{r}
# #loading night light rasters for feb 7
 feb705 <- raster(here::here("data", "VNP46A1", "VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
 feb706 <- raster(here::here("data", "VNP46A1", "VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
# #loading night light rasters for feb 16
feb1605 <- raster(here::here("data", "VNP46A1", "VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
feb1606 <- raster(here::here("data", "VNP46A1", "VNP46A1.A2021047.h08v06.001.2021048091105.tif"))
```

I now need to reclassify the difference raster, to assume that any location which experienced a drop of more than 200 nW cm-2sr-1 also experienced a blackout.

```{r}
t7_05  <- terra::rast(here::here("data","VNP46A1","VNP46A1.A2021038.h08v05.001.2021039064328.tif"))
t7_06  <- terra::rast(here::here("data","VNP46A1","VNP46A1.A2021038.h08v06.001.2021039064329.tif"))
t16_05 <- terra::rast(here::here("data","VNP46A1","VNP46A1.A2021047.h08v05.001.2021048091106.tif"))
t16_06 <- terra::rast(here::here("data","VNP46A1","VNP46A1.A2021047.h08v06.001.2021048091105.tif"))

r7  <- terra::merge(t7_05,  t7_06)
r16 <- terra::merge(t16_05, t16_06)

tmp7  <- tempfile(fileext = ".tif")
tmp16 <- tempfile(fileext = ".tif")
rast7  <- mosaic(feb706,  feb705,  fun = mean, filename = tmp7,  overwrite = TRUE)  # RasterLayer
rast16 <- mosaic(feb1606, feb1605, fun = mean, filename = tmp16, overwrite = TRUE)  # RasterLayer

if (!compareRaster(rast7, rast16, extent=TRUE, rowcol=TRUE, crs=TRUE, res=TRUE, stopiffalse=FALSE)) {
  rast16 <- projectRaster(rast16, rast7, method = "bilinear")
}

# 3) Difference + threshold (STILL raster, no terra objects here)
diff_bool <- (rast7 - rast16) > 200     # RasterLayer TRUE/FALSE
diff_bool[diff_bool == 0] <- NA    

mask_t   <- terra::rast(diff_bool)      # SpatRaster
diffvect <- terra::as.polygons(mask_t, dissolve=TRUE, na.rm=TRUE) |>
            sf::st_as_sf() |>
            sf::st_make_valid()

```

```{r}
# # check crs
# st_crs(rast16) # WGS84
# st_crs(rast7) # WGS84
# 
# # do the 2 extents match?
# if(ext(rast16) == ext(rast7)){
#   print("extents match")
# } else{
#   print("extents do not match")
# }
# 
# #conditional difference
# diff <- rast7 - rast16 > 200 #making a boolean
# #making those false's NA
# diff <- diff[diff == FALSE] <- NA
# class(diff)
# diff <- st_as_stars(diff)
```

Now vectorize! turning from a raster to a vector, to make a vector difference mask.

```{r}
 # diffvect <- st_as_sf(diff) %>%
 #  st_make_valid()
```

Then we crop

```{r}
#cropping to a bbox
# crop_bbox <-st_bbox(c(
#                      xmin = -96.5,
#                      ymin = 29,
#                      xmax = -94.50,
#                      ymax = 30.5),
# crs = st_crs(diffvect))
# #adding that as a layer
# crop <- st_as_sfc(crop_bbox)
# #cropping the dataset
# croppeddiff <- diffvect[crop,]
# st_bbox(croppeddiff)
# #changing the CRS
# croppeddiff <- croppeddiff %>%
#   st_transform('EPSG:3083')
#transforming
diff_ll <- st_transform(diffvect, 4326)

#cropping to a bbox
croppeddiff <- st_crop(
  diff_ll,
  xmin = -96.5, xmax = -94.50,
  ymin = 29.0,  ymax = 30.5
)

#transforming
croppeddiff <- st_transform(croppeddiff, 3083)

```

Excluding highways from the cropped blackout mask.

```{r}
#transforming into a crs with meters
transroad <- roads %>%
  st_transform('ESRI:102009')
roads <- roads %>%
  st_transform('ESRI:102009')
croppeddiff <- croppeddiff %>%
  st_transform('ESRI:102009')
#buffering for within 200 meters 
transroad <- st_buffer(transroad, 200)
#using St_union-kills my computer, doesn't work
#st_union(transroad,roads)
transroadbuff <- st_union(transroad)

```

identify the number of homes likely impacted by blackouts

```{r}

no_freeway <-st_difference(croppeddiff, transroadbuff)
tm_shape(no_freeway)+
  tm_polygons(col='green') +
  tm_basemap('OpenStreetMap')
```
# Maps 1 and 2
These maps illustrate the night light intensity before and after the first two storms
```{r}
tmap_mode("plot") 

# Changing everyone's CRS to match
r7_3083  <- terra::project(terra::rast(rast7),  "EPSG:3083", method = "bilinear")
r16_3083 <- terra::project(terra::rast(rast16), "EPSG:3083", method = "bilinear")
roads_3083       <- sf::st_transform(roads, 3083)
diffvect_3083    <- sf::st_transform(diffvect, 3083)
croppeddiff_3083 <- sf::st_transform(croppeddiff, 3083)
transroadbuff_3083 <- sf::st_transform(transroadbuff, 3083)

# making no highway data (but also the same CRS)
no_highway_3083 <- try(
  sf::st_difference(croppeddiff_3083, transroadbuff_3083),
  silent = TRUE
)
if (inherits(no_highway_3083, "try-error")) {
  touch <- lengths(sf::st_intersects(croppeddiff_3083, transroadbuff_3083)) > 0
  keep  <- croppeddiff_3083[!touch, , drop=FALSE]
  cut   <- sf::st_difference(croppeddiff_3083[touch, , drop=FALSE], transroadbuff_3083)
  cut   <- cut[!sf::st_is_empty(cut), , drop=FALSE]
  no_highway_3083 <- rbind(keep, cut)
}

# Cropping to match the BBOX
bb3083 <- sf::st_as_sfc(sf::st_bbox(croppeddiff_3083))
r7_3083   <- terra::crop(r7_3083,   terra::vect(bb3083))
r16_3083  <- terra::crop(r16_3083,  terra::vect(bb3083))

# Making the maps
m_before <- tm_shape(r7_3083) +
  tm_raster(style = "cont", palette = viridisLite::magma(9), title = "Radiance (Feb 7)") +
  tm_shape(roads_3083) + tm_lines(col = "grey30", lwd = 0.6) +
#  tm_basemap('OpenStreetMap')+
  tm_layout(frame = FALSE)

m_after <- tm_shape(r16_3083) +
  tm_raster(style = "cont", palette = viridisLite::magma(9), title = "Radiance (Feb 16)") +
  tm_shape(roads_3083) + tm_lines(col = "grey30", lwd = 0.6) +
#  tm_basemap('OpenStreetMap')+
  tm_layout(frame = FALSE)

m_mask <- tm_shape(croppeddiff_3083) +
  tm_polygons(col = "tomato", alpha = 0.6, title = "Blackout mask (>200)") +
  tm_shape(roads_3083) + tm_lines(col = "grey30", lwd = 0.6) +
  tm_basemap('OpenStreetMap')+
  tm_layout(frame = FALSE)

m_nohwy <- tm_shape(no_highway_3083) +
  tm_polygons(col = "forestgreen", alpha = 0.6, title = "Mask minus 200 m highways") +
  tm_shape(transroadbuff_3083) + tm_borders(col = "grey50", lwd = 0.6) +
  tm_shape(roads_3083) + tm_lines(col = "grey30", lwd = 0.6) +
  tm_basemap('OpenStreetMap')+
  tm_layout(frame = FALSE)

tmap_arrange(m_before, m_after, m_mask, m_nohwy, ncol = 2)

```
# Map of homes in Huston that lost power
```{r}
#filtering for homes that lost power
#subtracting R7-R16 (m_difference)
vis_bool <- (r16_3083 - r7_3083 ) < 200  
diff_3083 <- r16_3083 - r7_3083 

mm <- terra::global(vis_bool, c("min","max"), na.rm=TRUE)
maxabs <- max(abs(mm[1,1]), abs(mm[1,2]))
brks <- seq(-maxabs, maxabs, length.out = 11)

m_difference <- tm_shape(vis_bool) +
  tm_raster(
    title   = expression("Homes that lost power (Feb 16 - Feb 7), Measured by Change in Radiance"),
    breaks  = brks, style="pretty",
    palette = "RdPu"  
  ) +
  tm_shape(roads_3083) + tm_lines(col = "black", lwd = 0.6) +
  tm_layout(frame = FALSE)

mask_gt200_3083 <- terra::ifel(vis_bool > 200, 1, NA)

m_diff_mask <- m_difference +
  tm_shape(mask_gt200_3083) + tm_raster(alpha=0.35, palette=c("#000000"), legend.show=FALSE)

m_diff_mask
#mapping the difference
```
# Estimate of the number of homes in Huston that lost power
```{r}
#| message: false
#| warning: false

# Assuming 'r' is your SpatRaster object
# If you don't have a SpatRaster object, you can create one for testing
vis_bool <- rast(nrows=10, ncols=10, xmin=-180, ymin=-90)
values(vis_bool) <- 1:ncell(vis_bool)

# 1. Extract coordinates and values
coords <- crds(vis_bool)
vals <- values(vis_bool)

# 2. Combine into a data frame
vis_bool_df <- data.frame(X = coords[,1],
                 Y = coords[,2],
                 Value = vals)

# 3. Write to CSV
write.csv(vis_bool_df, "vis_bool_df.csv", row.names = FALSE)

vis_bool %>% 
  group_by(grade) %>% 
   summarise(Percent = round((n() / 8988) * 100, 1)) %>% #number of observations in the group, divided by total rows, multiplied by 100, and rounded to 1 decimal place
  mutate(Percent = paste0(Percent, "%")) %>% 
  rename(HOLC_Grade = grade) %>%   # capitalizing
  st_drop_geometry() %>% 
  # creating the table
  kbl(booktabs = T, caption = "Proportion of Census Block groups") %>%
  kable_styling(latex_options = "striped")
```
# (GG)Plot comparing the distributions of median household income for census tracts that did and did not experience blackouts
```{r}
```
# Result Summary